Sell Prevention and Buy Cure: An Exploration of Indirect Conflicts

### 2013-11-21 iced-2013/paper.tex
ICED Draft 4 done.

Awareness techniques have been proposed and studied to aid in developer
understanding, efficiency, and quality of software produced. Some of these techniques have focused 
on either \textit{direct} or 
\textit{indirect conflicts} in regards to prevent, detect, or resolve these conflicts as they arise
from a result of source code changes. While the techniques and tools for direct conflicts have had
large success, tools either proposed or studied for indirect conflicts have had common issues of
information overload, false positives, and scalability. To better understand these issues, 
we performed a mixed method study involving interviews with 19 developers from 12 commercial and open
cource projects, and a survey with 78 responses. Through grounded theory techniques, 
we found that: indirect conflicts are a real and frequent problem in
software development, developers prefer to use detection and resolution processes or tools
over those of prevention, developers do not want awareness mechanisms which provide non actionable results,
and that there exists a gap in software evolution analytical tools.

@@

While indirect conflict tools have shown potential from developer studies, some of the same problems continue
to arise throughout most, if not all tools. The most prevalent issue is that of false positives and information
overload. Through case studies, developers have noted that current indirect conflict tools provide too many 
false positive results leading to information overload and the tool eventually being
ignored~\cite{Sarma:2007:TSA, Servant:2010:CPI}. A second primary issue is that of dependency identification and
tracking. Many different dependencies have been proposed and used in indirect conflict tools such as method 
invocation~\cite{Trainer:2005:BGT}, and class signatures~\cite{Sarma:2007:TSA} with varying success, but the 
identification of failure inducing changes, other than those which are already identifiable by other means such
as compilers, and unit tests, to these dependencies still remains an issue. Dependency tracking issues are
also compounded by the scale of many software development projects leading to further information overload.
Lastly, social factors have also shown to cause issues as Cataldo et al.~\cite{Cataldo:2006:ICR} notion of socio-technical
congruence has been leveraged in several indirect conflict tools~\cite{Kwan:2011:ESC, Begel:2010:CDE, Borici:2012:CHA}.
Issues again of information overload, dependencies (of developer organizational structure), and scalability come up.

In the hopes of resolving some of the aforementioned issues surrounding indirect conflicts and their tools, we 
are interested in exploring the following three research questions:

\begin{description}
	\item[RQ1\namedlabel{itm:rq1}{RQ1}] \textit{What is the nature of indirect conflicts?}
	\item[RQ2\namedlabel{itm:rq2}{RQ2}] \textit{What types of process or tools are being used by developers in regards to indirect conflicts?}
	\item[RQ3\namedlabel{itm:rq3}{RQ3}] \textit{What do developers want from future indirect conflict tools?}
\end{description}

We interviewed 19 developers from across 12 commercial and open source projects, followed by a confirmatory survey of 78 
developers, as well as 5 confirmatory interviews, in order to answer the aforementioned questions. Our findings indicate that: 
indirect conflicts are very much a real 
and frequent problem in software development, that developers tend to prefer to use detection and resolution processes or tools
over those of prevention, that developers do not want awareness mechanisms which provide non actionable results, 
and that there exists a gap in software evolution analytical tools from the reliance on static analysis resulting in missed
context of indirect conflicts.

We continue this paper by first presenting our methodology used to perform this study in Section~\ref{sec:meth}. We
then present and discuss the results of our study in regards to our 3 research questions in Section~\ref{sec:results}
and Section~\ref{sec:disc}.
Next, an evaluation of how our study was performed in terms of the credibility
and quality of our results is described in Section~\ref{sec:eval}. We then cover related work in Section~\ref{sec:related}
and conclude in Section~\ref{sec:conc}.

@@

We performed a grounded theory study in two parts. First, a round of semi-structured interviews were conducted which 
addressed the 3 research questions as mentioned above. Second, a survey was conducted
which was used to confirm and test what was theorized from the interviews on a larger sample size as well as obtain
larger developer opinion of the subject.

\subsection{Grounded Theory}
We approached our study based on grounded theory as described by Corbin and Strauss~\cite{Corbin:1998:SP}.
Grounded theory is a qualitative research methodology that utilizes \textit{theoretical sampling} and
\textit{open coding} to create a theory ``grounded'' in the empirical data. For an exploratory study such as
the ours, grounded theory is well suited because it involves starting from very broad and abstract type questions, and
making refinements along the way as the study progresses and hypotheses begin to take shape. Grounded theory involves
realigning the sampling criteria throughout the course of the study to ensure that participants are able to answer new
questions that have been formulated in regards to forming hypotheses. In our study being presented, data collected from
both interviews and surveys (when open ended questions were involved) was analyzed using open and axial coding. Open coding involves
assigning codes to what participants said at a low sentence level or abstractly at a paragraph or full answer level. These
codes were defined as the study progressed and different hypotheses began to grow. We finally use axial coding it order to
link our defined codes through categories of grounded theory such as context and consequences. The ultimate goal of grounded 
theory is to produce a theory that is tied to the empirical data collected; our final theories can be seen in 
Section~\ref{sec:disc}. In Section~\ref{sec:eval}, we give a brief evaluation of our studying using 
3 criteria that are commonly used in evaluating grounded theory studies.

@@

\subsection{What is the nature of indirect conflicts?}

@@

\subsection{What types of process or tools are being used by developers in regards to indirect conflicts?}

@@

\textbf{Unwanted Awareness} As we have seen, indirect conflicts are found to be quite a serious problem that occur frequently,
sporadically, and differ greatly from case to case. These conditions pose large issues for the creation of generalizable theories
or tools in regards to indirect conflicts. These underlying complexities are the probable cause of disinterest of software developers
to proposed or implemented tools as discussed in Section~\ref{sec:related}. This inability to generalize is what we believe to be the
leading cause of information overload and false positives in the awareness system, causing developers to eventually ignore 
information being presented to them, rendering the system useless. These false positives are caused by a difference in what
developers consider to be false positives versus what awareness literature considers they are. This disjoint, as will
be seen, is caused by developers only considering events which cause some action to be taken on their part, to be true positives,
where as current awareness understanding would state any event which is related to an individual's 
work~\cite{Herbsleb06collaborationin}~\cite{Cataldo:2008:SCF}, to be true positives.

Developers have been found to have a great understanding of what they need to know about and more importantly what they don't in
their project awareness. To be able to fully understand a developer's awareness intuition, we can see from the results of this paper that
developers only want awareness of an event if the event forces the developer to take some action. In a sense, developers
don't care about what they don't need to account for. This sense of unwanted awareness is crucial to understand why generalized
awareness techniques of difficult to generalize problems, such as indirect conflicts through generalized changes to 
classes~\cite{Sarma:2007:TSA}, or methods~\cite{Trainer:2005:BGT,Servant:2010:CPI}, often encounter difficulties of false
positives and information overload. 
Developers simply do not want to know about events which effect them, but require no action on their part.

This unwanted awareness, or limited awareness, seems counter intuitive to current awareness understanding which would state that
being aware of all events in ones surrounding leads to higher productivity or other quality aspects. In theory this is correct,
but as it was found in practice, this is an incorrect assumption. Since software developers have limits on their time, awareness of
all events surrounding a developer's work or project is not possible. \textit{Developers prefer to spend their limited time dealing with
the awareness of events which cause them to take some action (changing code, communication, etc) rather than simply being aware
of events occurring around their work which pose no direct threat to the consumption of their time.}

@@

Two out of the three identified prevention methods taken by developers are simple blanket risk mitigation strategies
accomplished essentially by
not changing code (design by contract, and add and deprecate), while the third is simply developer experience and knowledge. Clearly,
developers are spending little to no time in prevention. Developers do however spend a large amount of time in detection and cure through the
writing of tests and the debugging of issues. In fact, most improvements mentioned by developers in regard to dealing with indirect
conflicts occurred at either the detection or cure levels. Obviously, developers either prefer to spend their time in curative
measures, or do not posses the proper tools to take better action in the preventative stages.

The lack of prevention process and tools being used is believed to be due to 2 factors. The first being the identification of dependencies.
Even with an experienced system architect, identifying dependencies and notifying those involved is a daunting task 
which is ignored more than dealt with. The second being the knowledge of when a dependency will fail also requires vast knowledge
of the product, more than anyone may have. This is compounded by the unique and sporadic nature of indirect conflicts. The
abundance of detection and curative process and tools on the other hand shows once again developers willingness to debug
real issues, that maybe even could have been prevented, rather than prevent the issues in the first place. With
this lack of prevention and abundance of curative measures, the question ultimately presented in this area is if 
curative measures are more productive than preventative measures.

@@

A last interesting observation about current curative measures being taken in industry, is that developers view testing of software 
as curative, when one could easily make the argument that it is preventative in that most tests are written to pass originally and
are kept in place to ensure future changes do not cause issues. This mindset may come from the notion that writing tests is originally
thought of as part of normal code writing, meaning that developers see the extra task of test coverage as part of feature implementation.
However, if a test never fails, could it not be said that it is preventing changes from causing it to fail rather than detecting
when failures or indirect conflicts do occur? This may suggest that if, given the write tools, developers may no longer view preventative
measures as a ``burden'' and may be more inclined to take a more balanced preventative approach rather than mostly curative.

@@

%mature point, static analysis
\textbf{Gap in Software Evolution Analysis} Indirect conflicts are more likely to occur,
from developer opinion, before a mature point is reached in the project's evolution. (This mature point cmay stem from a major release,
end of an iteration, a new feature being released or any point considered stable and reliable) However, this context of a 
mature point, as well as any other
potential contextual attributes, are rarely identified or accounted for in regards to indirect conflicts. Developers have said
that different change types may occur at different rates throughout a project's life time and that this
may drastically effect the outcomes of indirect conflict tools or processes.

@@

Relational database schemas are one of the highest sources of indirect
conflicts found in their projects, yet we know from Maule et al.~\cite{Maule:2008:IAD} that relational database schemas have 
been scarcely researched in terms of indirect conflicts. This falls in with our previous understanding of cross language
dependency tracking in that database schemas are independent of languages which may be on the receiving end of their output.

>>>>

Awareness techniques have been proposed and studied to aid developer
understanding, efficiency, and quality of software produced. Some of these techniques have focused 
on either \textit{direct} or 
\textit{indirect conflicts} in order to prevent, detect, or resolve these conflicts as they arise
from a result of source code changes. While the techniques and tools for direct conflicts have had
large success, tools either proposed or studied for indirect conflicts have had common issues of
information overload, false positives, and scalability. To better understand these issues, 
we performed a mixed method study involving interviews with 19 developers from 12 commercial and open
source projects, and a survey with 78 responses. Through grounded theory techniques, 
we found: events and conditions which indirect conflicts are likely to occur in
software development, developers prefer to use detection and resolution processes or tools
over those of prevention, developers do not want awareness mechanisms which provide non actionable results,
and that there exists a gap in software evolution analytical tools.

@@

While indirect conflict tools have shown potential from developer studies, some of the same problems continue
to arise throughout most, if not all tools. The most prevalent issue is that of false positives and information
overload. Through case studies, developers have noted that current indirect conflict tools provide too many 
false positive results leading to information overload and the tool eventually being
ignored~\cite{Sarma:2007:TSA, Servant:2010:CPI}. A second primary issue is that of dependency identification and
tracking. Many different dependencies have been proposed and used in indirect conflict tools such as method 
invocation~\cite{Trainer:2005:BGT}, and class signatures~\cite{Sarma:2007:TSA} with varying success, but the 
identification of failure inducing changes, other than those which are already identifiable by other means such
as compilers, and unit tests, to these dependencies still remains an issue. Dependency tracking issues are
also compounded by the scale of many software development projects leading to further information overload.
Lastly, social factors such as Cataldo et al.~\cite{Cataldo:2006:ICR} notion of socio-technical
congruence have been leveraged in indirect conflict tools~\cite{Kwan:2011:ESC, Begel:2010:CDE, Borici:2012:CHA}.
However, issues again of information overload, dependencies (in developer organizational structure), and scalability 
come up.

In order to fully understand the root causes of information overload, false positives, and
scalability issues in regards to indirect conflicts, we will take a step back and determine what events occur to
cause indirect conflicts, when they occur, and if conditions exist to provoke more of these events. 
We then set out to understand what mitigation strategies developers currently use as opposed to those created
by academia. Through this exploration, we look to find what can be accomplished moving forward with indirect conflicts
in both research and industry. To accomplish these goals, we ask the following:

\begin{description}
	\item[RQ1\namedlabel{itm:rq1}{RQ1}] \textit{What events or conditions lead to indirect conflicts?}
	\item[RQ2\namedlabel{itm:rq2}{RQ2}] \textit{What mitigation techniques are used by developers in regards to indirect conflicts?}
	\item[RQ3\namedlabel{itm:rq3}{RQ3}] \textit{What do developers want from future indirect conflict tools?}
\end{description}

We interviewed 19 developers from across 12 commercial and open source projects, followed by a confirmatory survey of 78 
developers, and 5 confirmatory interviews, in order to answer the aforementioned questions. Our findings indicate that: 
indirect conflicts occur frequently and are likely caused by software contract changes and a lack of understanding,
developers tend to prefer to use detection and resolution processes or tools
over those of prevention, developers do not want awareness mechanisms which provide non actionable results, 
and there exists a gap in software evolution analytical tools from the reliance on static analysis resulting in missed
context of indirect conflicts.

@@

We performed a a mixed method study in three parts. First, a round of semi-structured interviews were conducted which 
addressed the 3 research questions as mentioned above. Second, a survey was conducted
which was used to confirm and test what was theorized from the interviews on a larger sample size as well as obtain
larger developer opinion of the subject. Third, 5 confirmatory interviews were conducted by re-interviewing original
interview participants to once again confirm and test what had been learned and the results derived by the paper.
Grounded theory techniques were used to analyze the information provided from all three data gathering processes.

\subsection{Grounded Theory}
We approached our study based on grounded theory as described by Corbin and Strauss~\cite{Corbin:1998:SP}.
Grounded theory is a qualitative research methodology that utilizes \textit{theoretical sampling} and
\textit{open coding} to create a theory ``grounded'' in the empirical data. For an exploratory study such as
the ours, grounded theory is well suited because it involves starting from very broad and abstract type questions, and
making refinements along the way as the study progresses and hypotheses begin to take shape. Grounded theory involves
realigning the sampling criteria throughout the course of the study to ensure that participants are able to answer new
questions that have been formulated in regards to forming hypotheses. In our study being presented, data collected from
both interviews and surveys (when open ended questions were involved) was analyzed using open and axial coding. Open coding involves
assigning codes to what participants said at a low sentence level or abstractly at a paragraph or full answer level. These
codes were defined as the study progressed and different hypotheses began to grow. We finally use axial coding it order to
link our defined codes through categories of grounded theory such as context and consequences. 
In Section~\ref{sec:eval}, we give a brief evaluation of our studying using 
3 criteria that are commonly used in evaluating grounded theory studies.

@@

\subsection{What events or conditions lead to indirect conflicts?}

@@

\subsection{What mitigation techniques are used by developers in regards to indirect conflicts?}

@@

\textbf{Unwanted Awareness} As we have seen, indirect conflicts are found to be quite a serious problem that occur frequently,
sporadically, and differ greatly from case to case. These conditions pose large issues for the creation of generalizable theories
or tools in regards to indirect conflicts. These underlying complexities are the probable cause of disinterest of software developers
to proposed or implemented tools as discussed in Section~\ref{sec:related}. This inability to generalize is what we believe to be the
leading cause of information overload and false positives in the awareness system, causing developers to eventually ignore 
information being presented to them, rendering the system useless. These false positives are caused by a difference in what
developers consider to be false positives versus what awareness literature considers they are. This disjoint, as will
be seen, is caused by developers only considering events which cause some action to be taken on their part, to be true positives,
where as current awareness understanding would state any event which is related to an individual's 
work~\cite{Herbsleb06collaborationin}~\cite{Cataldo:2008:SCF}, to be true positives.
``You need a good understanding of what the code change is or else you will have a lot of false positives'' said one developer,
showing that not all changes around an object should be reported for awareness.

Developers have been found to have a great understanding of what they need to know about and more importantly what they don't in
their project awareness. To be able to fully understand a developer's awareness intuition, we can see from the results of this paper that
developers only want awareness of an event if the event forces the developer to take some action. In a sense, developers
don't care about what they don't need to account for. ``We would want a high probability of the [reported] change being an issue'', meaning
that the developer only wants the awareness if the item will require action on their part to resolve the issue.
This sense of unwanted or limited awareness is crucial to understand why generalized
awareness techniques of difficult to generalize problems, such as indirect conflicts through generalized changes to 
classes~\cite{Sarma:2007:TSA}, or methods~\cite{Trainer:2005:BGT,Servant:2010:CPI}, often encounter difficulties of false
positives and information overload. 
Developers simply do not want to know about events which effect them, but require no action on their part.

This unwanted awareness, or limited awareness, seems counter intuitive to current awareness understanding which would state that
being aware of all events in ones surrounding leads to higher productivity or other quality aspects. In theory this is correct,
but as it was found in practice, this is an incorrect assumption. 
In regards to this full awareness, one developer said ``There is no room for this in [our company], as tools are already in place for 
analysis of change[s] and code review takes care of the rest''. Since software developers have limits on their time, awareness of
all events surrounding a developer's work or project is not possible. \textit{Developers prefer to spend their limited time dealing with
the awareness of events which cause them to take some action (changing code, communication, etc) rather than simply being aware
of events occurring around their work which pose no direct threat to the consumption of their time.}

@@

Two out of the three identified prevention methods taken by developers are simple blanket risk mitigation strategies
accomplished essentially by
not changing code (design by contract, and add and deprecate), while the third is simply developer experience and knowledge. Clearly,
developers are spending little to no time in prevention. Developers do however spend a large amount of time in detection and cure through the
writing of tests and the debugging of issues. In fact, most improvements mentioned by developers in regard to dealing with indirect
conflicts occurred at either the detection or cure levels. Obviously, developers either prefer to spend their time in curative
measures, or do not posses the proper tools to take better action in the preventative stages. ``You're reaction time is much more crucial''
said one developer in that resolution tools are believed to be or larger importance as once issues have occurred, it doesn't
matter what prevention was taken, the issue must simply be resolves as quick as possible now.

The lack of prevention process and tools being used is believed to be due to 2 factors. The first being the identification of dependencies.
Even with an experienced system architect, identifying dependencies and notifying those involved is a daunting task 
which is ignored more than dealt with. The second being the knowledge of when a dependency will fail, also requires vast knowledge
of the product, more than anyone may have. This is compounded by the unique and sporadic nature of indirect conflicts. These 2 factors
are what ultimately have led to the amount of false positives and information overload seen in previous tools. The
abundance of detection and curative process and tools on the other hand shows once again developers willingness to debug
real issues, that maybe even could have been prevented, rather than prevent the issues in the first place. With
this lack of prevention and abundance of curative measures, the question ultimately presented in this area is if 
curative measures are more productive than preventative measures.

@@

A last interesting observation about current curative measures being taken in industry, is that developers view testing of software 
as curative, when one could easily make the argument that it is preventative in that most tests are written to pass originally and
are kept in place to ensure future changes do not cause issues. This mindset may come from the notion that writing tests is originally
thought of as part of normal code writing, meaning that developers see the extra task of test coverage as part of feature implementation.
However, if a test never fails, could it not be said that it is preventing changes from causing it to fail rather than detecting
when failures or indirect conflicts do occur? This may suggest that if, given the write tools, developers may no longer view preventative
measures as a ``burden'' and may be more inclined to take a more balanced preventative approach rather than mostly curative, as
``prevention is the goal'' was commonly said by developers.

@@

%mature point, static analysis
\textbf{Gap in Software Evolution Analysis} Indirect conflicts are more likely to occur,
from developer opinion, before a mature point is reached in the project's evolution. ``Once you get the API stable, people are 
better at communicating changes in regards to dependency concerns.''(This mature point may stem from a major release,
end of an iteration, a new feature being released or any point considered stable and reliable) However, this context of a 
mature point, as well as any other
potential contextual attributes, are rarely identified or accounted for in regards to indirect conflicts. Developers have said
that different change types may occur at different rates throughout a project's life time and that this
may drastically effect the outcomes of indirect conflict tools or processes.

@@

Relational database schemas are one of the highest sources of indirect
conflicts found in their projects, ``it breaks stuff all over the place'',
yet we know from Maule et al.~\cite{Maule:2008:IAD} that relational database schemas have 
been scarcely researched in terms of indirect conflicts. This falls in with our previous understanding of cross language
dependency tracking in that database schemas are independent of languages which may be on the receiving end of their output.
